Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-08-05 Delaneyetal2024.tex
small changes

The second experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, and friction factors.
Additionally, we examine the effects differing water discharge smoothing periods from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their daily averages. 
This creates a time series that is detrended from he seasonal variations in the model outputs and has an approximately normal distribution.
We then present the variability in the model output by taking the standard deviation of this detrended time series.

@@

Across the range of parameters examined, variability in all model outputs (i.e. velocity, shear stress, and width-integrated shear stress) remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux and thus subglacial channel size in \icesheet{} drives the larger width integrated shear stress variations (Figures~\ref{fig:Qw}~and~\ref{fig:Qw}\,c and f).

@@

Here, we evaluate the relative sensitivity of different channel types as they scale with different water discharge, channel shape and hydraulic gradient.
These include subaerial channels (Figure~\ref{fig:chan_types} c), steady-state R-channels \citep[Figure~\ref{fig:chan_types} d]{rothlisberger1972} and pipe-flow (i.e. R-channels of fixed size that do not adjust their size to discharge conditions; Figure~\ref{fig:chan_types} a).
These formulations do not allow us to evaluate non-steady state R-channels (Section~\ref{sect:sub_mode}, Figure~\ref{fig:chan_types} b).
The sediment transport capacity is calculated with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

@@

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
The results of the 12 combinations of shear stress and sediment transport capacity are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

@@

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different channel types and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow channel type.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$ for the MPM (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

The manuscript's first objective is to establish if water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an discharge event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most channels.
This incoherent relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Water discharge and sediment transport capacity covariance could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case.

>>>>

The second experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, and friction factors.
Additionally, we examine the effects of differing water discharge smoothing periods from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their daily averages. 
This creates a time series that is detrended from the seasonal variations in the model outputs and has an approximately normal distribution.
We then present the variability in the model output by taking the standard deviation of this detrended time series.

@@

Across the range of parameters examined, variability in all model outputs (i.e. velocity, shear stress, and width-integrated shear stress) remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects of the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux and thus subglacial channel size in \icesheet{} drives the larger width integrated shear stress variations (Figures~\ref{fig:Qw}~and~\ref{fig:Qw}\,c and f).

@@

Here, we evaluate the relative sensitivity of different channel types as they scale with water discharge, channel shape, and hydraulic gradient.
These include subaerial channels (Figure~\ref{fig:chan_types} c), steady-state R-channels \citep[Figure~\ref{fig:chan_types} d]{rothlisberger1972} and pipe-flow (i.e. R-channels of fixed size that do not adjust their size to discharge conditions; Figure~\ref{fig:chan_types} a).
These formulations do not allow us to evaluate non-steady state R-channels (Section~\ref{sect:sub_mode}, Figure~\ref{fig:chan_types} b).
The sediment transport capacity is calculated with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

@@

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results are summarised in Table~\ref{tab:eqs1}.
The results of the 12 combinations of shear stress and sediment transport capacity are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

@@

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different channel types and  $\alpha$ values.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow channel type.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$ for the MPM (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

The manuscript's first objective is to establish if water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during a discharge event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most channels.
This incoherent relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Water discharge and sediment transport capacity covariance could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case.
